```{r}
setwd("C:/Users/jiang/OneDrive/Documents/R/finalProject4030/Forest-Fires-Data-Set-Portugal")

library(car)
library(lmtest)
library(skedastic)
library(sandwich)
```

```{r}
df = read.table("fire.txt", header = TRUE)

df_trans = df

df_trans$month = factor(df_trans$month)
df_trans$day = factor(df_trans$day)
df_trans$log_area = log(df_trans$area)

```

```{r}
df_inter = df_trans[, c(
  "log_area", "day", "month",
  "X", "Y", "FFMC", "DMC", "DC", "ISI", "temp", "RH", "wind"
)]

summary(df_inter)
```

First, we test the interaction of day with other covariates, expect month. We have 200 degrees of freedom in this model.

```{r}
fit_day_inter = lm(
  log_area ~ day * (X + Y + FFMC + DMC + DC + ISI + temp + RH + wind),
  data = df_inter
)

summary(fit_day_inter)
anova(fit_day_inter)
```

We don't expect anything to be significant, as it is very hard to explain in real life, but we follow the steps first.
Here, at the first few steps, we find that day:DMC is still close to be significant, so we don't think is proper to drop `DMC`.
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:FFMC)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:temp)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - temp)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:RH)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - FFMC)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:DC)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:wind)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - DC)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:Y)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - Y)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:X)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - X)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:ISI)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - RH)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - wind)
anova(fit_day_inter)
```
If we set the significant level as 0.05, we will remove all covariates. 
And if we set alpha=1, we will reach to 4 left: `day`,`DMC`.`ISI`,`day:DMC`, and last two are significant,
though this interaction effect is very hard to explain.

We also test interaction effects of month, yet nothing is significant.