---
title: "Fire"
author: "Xinlu Jiang"
date: "2026-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("C:/Users/jiang/OneDrive/Documents/R/finalProject4030/Forest-Fires-Data-Set-Portugal")
library(car)
library(lmtest)
library(skedastic)
library(sandwich)
```

## Introduction to "Fire"

This data set contains measurements on 270 forest fires in the northeast region of Portugal.

```{r}
df = read.table("fire.txt", header = TRUE)
summary(df)
```

Our response variable is `area`, by our first eye, we noticed that it has a median of 6.37,
 a mean of 24.6, while the 3rd qu of 15.42, indicating that it does not follow normal 
 distribution for sure and we might transform it into log form in the future.

```{r}
hist(df$area)
```

From our observation of the plot, we are sure that the area probably needs a log transformation.
 And because it is strictly larger than 0, we can just use `log(area)` instead of `log(area+1)`,
 which is what we learned on class.

Quantitative covariates includes `FFMC`, `DMC`, `DC`, `ISI`, `temp`, `RH`, `wind`, and 
 location information `X` and `Y`. While the rest, `month` and `day` are qualitative 
 variables that will be used later for fixed/random effects.

### Exploratory Simple Linear Regressions

This part is not required in guidance, yet we find it might be useful as we goes further.

```{r}
df$log_area = log(df$area)
```

```{r}
quant_vars = c("X", "Y", "FFMC", "DMC", "DC", "ISI", "temp", "RH", "wind")

uni_models = lapply(quant_vars, function(v) {
  lm(as.formula(paste("log_area ~", v)), data = df)
})

names(uni_models) = quant_vars
```

```{r}
uni_results = data.frame(
  variable = quant_vars,
  coefficient = sapply(uni_models, function(m) coef(m)[2]),
  p_value = sapply(uni_models, function(m) summary(m)$coefficients[2, 4]),
  r_squared = sapply(uni_models, function(m) summary(m)$r.squared),
  adj_r_squared = sapply(uni_models, function(m) summary(m)$adj.r.squared)
)

uni_results
```

From the chart, we find that none of the covariates on its own is statistically significant
 (alpha=0.05).  Most predictors have extremely small \(R^2\) values, indicating that each 
 variable alone explains almost none of the variation in burned area. The only variable 
 close to marginal significance is `ISI`, with a p-value around 0.0705, but it is still
 not significant. And `ISI` itself is actually a combination of `wind` and other indicators.

### Categorical Exploratory Analysis

We also examine the categorical variables `month` and `day`. Instead of immediately fitting
 a regression model with dummy variables, we first calculate the group-level sample size and
  average burned area. This helps us understand whether certain months or weekdays are 
  associated with larger observed fire areas.

```{r}
# to get the right sequence
df$month = factor(
  df$month,
  levels = c("jan", "feb", "mar", "apr", "may", "jun",
             "jul", "aug", "sep", "oct", "nov", "dec")
)

df$day = factor(
  df$day,
  levels = c("mon", "tue", "wed", "thu", "fri", "sat", "sun")
)
```

```{r}
month_counts = table(df$month)

barplot(
  month_counts,
  main = "Frequency of Observations by Month",
  xlab = "Month",
  ylab = "Count"
)
```

From this plot, we can see that Auguest and September is the peak season of forest
 fire, which aligns to our guess. 
```{r}
day_counts = table(df$day)

barplot(
  day_counts,
  main = "Frequency of Observations by Day",
  xlab = "Day",
  ylab = "Count"
)
```

Though might not be significant, we can see that weekends, including Friday have a
 higher chance of fire. We guess that this might be related to human activities,
 as tourist might be a significant reason of forest fire.
```{r}
boxplot(
  log_area ~ month,
  data = df,
  xlab = "Month",
  ylab = "log(area)",
  main = "Distribution of log(area) by Month"
)
```

```{r}
boxplot(
  log_area ~ day,
  data = df,
  xlab = "Day",
  ylab = "log(area)",
  main = "Distribution of log(area) by Day"
)
```

Different from our observation of frequence, the distribution of area by month
 seems to be quite average. But we can see some trends that weekends seem to have
 larger, fire area than weekdays.

## Basic Model

### OLS and tests

```{r}
# We preserve only quantitative covariates at this stage

df_quant = subset(df, select = -c(month, day, log_area)) # we cannot use transformation here

ols_model = lm(area ~ ., data = df_quant)
summary(ols_model)
```

We notice that none of the covariates is statistically significant at the 5% level
 at this stage, and the adjusted \(R^2\) is very low.

Our required linear constrains to test is 
$$
H_0:50\beta_{DHC}+10\beta_{RH}=3\beta_{lSl}+12\beta_{temp}=3.
$$

```{r}
constraints <- c("50*DC + 10*RH = 3", 
                 "6*ISI + 12*temp = 3")
f_linear_constrain_test <- linearHypothesis(ols_model, constraints, test = "F")
print(f_linear_constrain_test)

wald_linear_constrain_test <- linearHypothesis(ols_model, constraints, test = "Chisq")
print(wald_linear_constrain_test)
```

The exact F test gives a p-value of 0.1527, and the asymptotic Wald test gives
 a p-value of 0.1506. Both tests lead to the same conclusion: we fail to reject 
 the null hypothesis at the 5% significance level. In other words, the data do 
 not provide sufficient evidence against the two proposed linear constraints.

The chi-square statistic is approximately twice the F statistic because the Wald
 statistic uses the number of restrictions as the scaling factor. In this case, 
 there are \(q = 2\) restrictions, so \(3.786 \approx 2 \times 1.893\). The small 
 difference in p-values comes from the fact that the exact F test uses an 
 \(F_{2,260}\) reference distribution, while the Wald test uses an asymptotic 
 \(\chi^2_2\) reference distribution.

The score test and likelihood ratio test are also commonly used asymptotic tests 
for model restrictions. Conceptually, the Wald test evaluates how far the 
unrestricted estimates are from satisfying the null constraints, while the score 
test evaluates the slope of the likelihood at the restricted estimate. The 
likelihood ratio test compares the fit of the unrestricted and restricted 
models directly.

## Heteroscedasticity

OLS standard errors rely on the homoscedasticity assumption. Therefore, we first examine
 the residual-versus-fitted plot and then use formal heteroscedasticity tests to verify 
 the visual evidence.
```{r}
plot(ols_model, which = 1)
```

The residual-versus-fitted plot shows that the basic OLS model with the original response
 variable `area` is not ideal. Most residuals are concentrated near zero, while a few 
 observations have extremely large positive residuals. In addition, the fitted values 
 include negative values, which is not reasonable for burned area because `area` is 
 nonnegative by definition. So we would like to take a look of transformed model first.

```{r}
trans_model = lm(log(area)~ ., df_quant)
plot(trans_model, which = 1)
```

```{r}
bptest(ols_model)
bptest(trans_model)
```

We then apply the studentized Breusch-Pagan test to both the original OLS model and the transformed model.

For the original model, the p-value is 0.7981, so we fail to reject the null hypothesis of homoscedasticity.
 For the transformed model, the p-value is 0.1091, which is still larger than 0.05. Therefore, we also fail 
 to reject the null hypothesis of homoscedasticity for the transformed model.

In conclusion, the Breusch-Pagan test does not provide significant evidence of heteroscedasticity in either 
model. Therefore, there is no strong statistical evidence that heteroscedasticity is the main issue in this 
stage. However, the residual plots still show outliers and a less satisfactory residual pattern for the 
original response model, so we continue to report robust standard errors as a sensitivity check.

### Robust Standard Errors

```{r}
se_ols <- sqrt(diag(vcov(ols_model)))
se_hc3 <- sqrt(diag(vcovHC(ols_model, type = "HC3")))

compare_se <- data.frame(
  Variable = names(se_ols),
  OLS_SE = se_ols,
  HC3_SE = se_hc3
  )

print(compare_se)
```
In this model, the HC3 robust standard errors are smaller than the conventional OLS standard errors for most coefficients.

```{r}
robust_t_test <- coeftest(ols_model, vcov = vcovHC(ols_model, type = "HC3"))
print(robust_t_test)
summary(ols_model)
```

The robust t test shows a even better result, though none of the covariates are statistical significant.

```{r}
robust_ci <- coefci(ols_model, vcov = vcovHC(ols_model, type = "HC3"), level = 0.95)
print("---hc3 vcov---")
print(robust_ci)
ols_ci <- confint(ols_model, level = 0.95)
print("---ols(hc0) vcov---")
print(ols_ci)
```

Similarly, the 95% robust CIs are smaller than the CIs from OLS.

### Iteratively Reweighted Least Squares

```{r}
# IRLS with residual variance modeled as a function of temp

irls_model = ols_model
residuals_irls = irls_model$residuals

w = rep(1, nrow(df_quant))
err = Inf
tol = 1e-10
niter = 0
maxiter = 1e3

while (err > tol & niter < maxiter)
{
  
  aux = lm(log(residuals_irls^2) ~ log(temp), df_quant)
  err = sum(abs(w - exp(-aux$fitted.values))) / sum(w)
  w = exp(-aux$fitted.values)
  WLS = lm(area ~ ., df_quant, weights = w)
  residuals_irls = WLS$residuals
  niter = niter + 1
  
}

summary(aux)
summary(WLS)
```

```{r}
gamma_hat = coef(aux_model)

alpha_0_hat = exp(gamma_hat[1])
alpha_1_hat = gamma_hat[2]

alpha_0_hat
alpha_1_hat

cat(
  paste0(
    "Estimated residual variance equation: sigma_i^2 = ",
    round(alpha_0_hat, 4),
    " * temp_i^",
    round(alpha_1_hat, 4)
  )
)
```

```{r}
ols_summary = summary(ols_model)
irls_summary = summary(irls_model)

compare_ols_irls = data.frame(
  Variable = names(coef(ols_model)),
  OLS_Estimate = coef(ols_model),
  OLS_SE = ols_summary$coefficients[, "Std. Error"],
  OLS_t = ols_summary$coefficients[, "t value"],
  OLS_p_value = ols_summary$coefficients[, "Pr(>|t|)"],
  IRLS_Estimate = coef(irls_model),
  IRLS_SE = irls_summary$coefficients[, "Std. Error"],
  IRLS_t = irls_summary$coefficients[, "t value"],
  IRLS_p_value = irls_summary$coefficients[, "Pr(>|t|)"]
)

print(compare_ols_irls)
```

```{r}
ols_ci = confint(ols_model, level = 0.95)
irls_ci = confint(irls_model, level = 0.95)

print("--- OLS 95% CI ---")
print(ols_ci)

print("--- IRLS/WLS 95% CI ---")
print(irls_ci)
```

```{r}
summary(w)
```

Compared with OLS, the IRLS estimates change substantially for several coefficients.
 For example, the coefficient of `temp` decreases from 1.1524 under OLS to 0.1311 
 under IRLS, and the coefficient of `wind` changes from 2.5191 to -0.1492. The standard
 errors are also generally smaller under IRLS. However, the inferential conclusion remains
the same: none of the covariates is statistically significant at the 5% level.

The confidence intervals from IRLS are generally narrower than the OLS confidence intervals.
 For example, the 95% CI for `temp` changes from \([-1.6071, 3.9118]\) under OLS to 
 \([-0.9781, 1.2402]\) under IRLS. Similarly, the 95% CI for `wind` changes from 
 \([-3.6896, 8.7279]\) under OLS to \([-2.5885, 2.2900]\) under IRLS. Even though 
 the IRLS intervals are narrower, they still contain zero for every covariate.