---
title: "Fire"
author: "Xinlu Jiang"
date: "2026-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(car);library(lmtest);library(skedastic);library(sandwich);
setwd("C:/Users/jiang/OneDrive/Documents/R/finalProject4030/Forest-Fires-Data-Set-Portugal")
```

## Introduction to "Fire"

This data set contains measurements on 270 forest fires in the northeast region of Portugal.

```{r}
df = read.table("fire.txt", header = TRUE)
summary(df)
```

Our response variable is `area`, by our first eye, we noticed that it has a median of 6.37, a mean of 24.6, while the 3rd qu of 15.42, indicating that it does not follow normal distribution for sure and we might transform it into log form in the future.

```{r}
hist(df$area)
```

Quantitative covariates includes `FFMC`, `DMC`, `DC`, `ISI`, `temp`, `RH`, `wind`, and location information `X` and `Y`. While the rest, `month` and `day` are qualitative variables that will be used later for fixed/random effects.

## Basic Model

### OLS and tests

```{r}
# We preserve quantitative covirates at this stage
df_quant = subset(df, select = -c(month,day))
ols_model = lm(area ~ ., df_quant)
summary(ols_model)
```

We notice that no covariates shows statistical significance at this stage, and the adjusted R-squared is very low.

Our required linear constrain to test is 
$$
H_0:50\beta_{DHC}+10\beta_{RH}=6\beta_{lSl}+12\beta_{temp}=3.
$$

```{r}
constraints <- c("50*DC + 10*RH = 3", 
                 "6*ISI + 12*temp = 3")
f_test <- linearHypothesis(ols_model, constraints, test = "F")
print(f_test)

wald_test <- linearHypothesis(ols_model, constraints, test = "Chisq")
print(wald_test)
```

The results from Exact F test and Asymptomatic Wald test agree each other and show 
 that we cannot reject the null hypothesis. The Chisq statistic is about two times of F 
 statistic, as the two has a difference of 2 degrees of freedom. Score test and LR test 
 are both asymptomatic test, and similar to Wald test. The only difference between Wald
 test and Score test is the variance they use.

### Heteroscedasticity

## Heteroscedasticity

OLS standard errors rely on the homoscedasticity assumption. Therefore, we first examine
 the residual-versus-fitted plot and then use formal heteroscedasticity tests to verify 
 the visual evidence.
```{r}
plot(ols_model, which = 1)
```

From the plot, we are surprised to find that there is no significant heteroscedasticity. We use test to verify our observation from the plot.

```{r}
bp_test_result = bptest(ols_model)
print(bp_test_result)
```

```{r}
squared_residuals <- resid(ols_model)^2

df_quant2 = subset(df_quant, select = -area)
aux_model <- lm(squared_residuals ~ .^2 + I(as.matrix(df_quant2)^2), data = df_quant2)
# We could also include interaction here if needed
r_sq <- summary(aux_model)$r.squared
n <- nrow(df_quant2)
lm_stat <- n * r_sq
p_val <- pchisq(lm_stat, df = aux_model$rank - 1, lower.tail = FALSE)
print(p_val)
```

```{r}
white(ols_model)
```

We could not reject the null hypothesis, thus we have conclude that the residuals are homoscedastic. However, we doubt the result of both tests, because we think most of our respond data are around 0, it is a serious right skewed data. And because BP test rely heavily on normality assumption, it may not work in our basic model. Though White test does not rely on normality assumption under large sample size, our model has about 200 degrees of freedom, might not enough for it to provide the correct result.

After all, at this stage, the variance of the random error terms non-constant is not a function of any of the quantitative covariates.

## Robust Standard Errors 

```{r}
se_ols <- sqrt(diag(vcov(ols_model)))
se_hc3 <- sqrt(diag(vcovHC(ols_model, type = "HC3")))

compare_se <- data.frame(
  Variable = names(se_ols),
  OLS_SE = se_ols,
  HC3_SE = se_hc3
  )

print(compare_se)
```

Different from what we learned, the HC3 SE is even smaller than OLS SE.

```{r}
robust_t_test <- coeftest(ols_model, vcov = vcovHC(ols_model, type = "HC3"))
print(robust_t_test)
summary(ols_model)
```

The robust t test shows a even better result, though non of the covariates are statistical significant.

```{r}
robust_ci <- coefci(ols_model, vcov = vcovHC(ols_model, type = "HC3"), level = 0.95)
print("---hc3 vcov---")
print(robust_ci)
ols_ci <- confint(ols_model, level = 0.95)
print("---ols(hc0) vcov---")
print(ols_ci)
```

Similarly, the 95% robust CIs are smaller than the CIs from OLS.

### Iteratively Reweighted Least Squares

```{r}

```

## Bootstrap

### Jackknife

```{r}
n = dim(df_quant)[1]
p = dim(df_quant)[2]

beta = ols_model$coefficients
SE_ols = summary(ols_model)$coefficients[, 2]

# Jackknife
betajack = matrix(0, n, p)
for (i in 1:n)
{
  jack = lm(area ~ ., df_quant[-i, ])
  betajack[i,] = jack$coefficients
}

SE_jack = apply(betajack, 2, sd) * sqrt((n - 1)^2 / n)

# Bootstrap
nboot = 10000
betaboot = matrix(0, nboot, p)
for (k in 1:nboot)
{
  boot = lm(area ~ ., df_quant[sample(n, replace = TRUE), ])
  betaboot[k,] = boot$coefficients
}
betabar_boot = colMeans(betaboot)
# Calculate Bootstrap SE
SE_boot = apply(betaboot, 2, sd)

se_comparison <- data.frame(
  Variable = names(SE_ols),
  OLS_SE = SE_ols,
  Jackknife_SE = SE_jack,
  Bootstrap_SE = SE_boot
)

print(se_comparison)
```

For most covirates, OLS_SE is larger than the SE of jackknife and bootstrap, while the latter two is about the same.

```{r}
alpha = 0.05

ci_ols <- confint(ols_model, level = 1 - alpha)
CInormal = cbind(betabar_boot - qnorm(1 - alpha / 2) * SE_boot, 
                 betabar_boot + qnorm(1 - alpha / 2) * SE_boot)
colnames(CInormal) = c("2.5%", "97.5%")
CIpercentile = cbind(apply(betaboot, 2, quantile, probs = alpha / 2), 
                     apply(betaboot, 2, quantile, probs = 1 - alpha / 2))
colnames(CIpercentile) = c("2.5%", "97.5%")
CIpivotal = cbind(2 * beta - apply(betaboot, 2, quantile, probs = 1 - alpha / 2), 
                  2 * beta - apply(betaboot, 2, quantile, probs = alpha / 2))
colnames(CIpivotal) = c("2.5%", "97.5%")
CI_wide <- data.frame(
  Variable = names(beta),
  OLS = sprintf("[%0.2f, %0.2f]", ci_ols[,1], ci_ols[,2]),
  Normal = sprintf("[%0.2f, %0.2f]", CInormal[,1], CInormal[,2]),
  Percentile = sprintf("[%0.2f, %0.2f]", CIpercentile[,1], CIpercentile[,2]),
  Pivotal = sprintf("[%0.2f, %0.2f]", CIpivotal[,1], CIpivotal[,2])
)

print(CI_wide)
```

The normal CI is smaller then OLS CI, which is different from our assumption. But we can see that the percentile CI is mostly smaller than the normal CI, while the Pivotal CI is exact the size of Percentile CI. The Pivotal CI can adjust itself according to the skewed distribution of the covariates.

### Permutation test of model and `Wind`

```{r}
set.seed(123)
B <- 10000
f_observed <- summary(ols_model)$fstatistic[1]
df_f1 <- summary(ols_model)$fstatistic[2]
df_f2 <- summary(ols_model)$fstatistic[3]
f_permuted <- numeric(B)

for (i in 1:B) {
  perm_y <- sample(df_quant$area)
  perm_model <- lm(perm_y ~ ., data = df_quant[, -which(names(df_quant) == "area")])
  f_permuted[i] <- summary(perm_model)$fstatistic[1]
}

p_val_F_precise <- (1 + sum(f_permuted >= f_observed)) / (1 + B)
print(p_val_F_precise)
```

```{r}
par(mar = c(5, 5, 4, 2) + 0.1, mgp = c(3, 1, 0))

hist(f_permuted, "FD", freq = FALSE, main = NA, 
     xlab = "Permutation F Test Statistics", ylab = "Density",
     col = "lightgray", border = "white")
curve(df(x, df_f1, df_f2), add = TRUE, lty = 2, lwd = 2, col = 1)
abline(v = qf(1 - alpha, df_f1, df_f2), col = 2, lty = 2, lwd = 4)
abline(v = quantile(f_permuted, 1 - alpha), col = 4, lty = 2, lwd = 4)
abline(v = f_observed, col = 3, lty = 2, lwd = 4)

legend("topright", 
       c("Theoretical Distribution", "Theoretical Quantile", 
         "Empirical Quantile", "Observed Statistic"),
       col = c(1, 2, 4, 3), lty = rep(2, 4), lwd = rep(4, 4), 
       cex = 0.8, bg = "white")

mtext(text = paste("Observed F:", round(f_observed, 2), 
                   "   Estimated P-value:", round(p_val_F_precise, 5)), 
      side = 1, line = -1.5, adj = 0.5, outer = FALSE, cex = 0.9, col = "black")
```

```{r}
model_reduced <- lm(area ~ . - wind, data = df_quant)
y_hat_reduced <- fitted(model_reduced)
res_reduced <- residuals(model_reduced)
t_observed <- summary(ols_model)$coefficients["wind", "t value"]
df_t <- summary(ols_model)$df[2]
t_permuted_res <- numeric(B)

for (i in 1:B) {
  res_star <- sample(res_reduced)
  y_star <- y_hat_reduced + res_star
  temp_data <- df_quant
  temp_data$area <- y_star
  model_star <- lm(area ~ ., data = temp_data)
  t_permuted_res[i] <- summary(model_star)$coefficients["wind", "t value"]
}

p_val_t_wind_precise <- (1 + sum(abs(t_permuted_res) >= abs(t_observed))) / (1 + B)
print(p_val_t_wind_precise)
```

```{r}
par(mar = c(5, 5, 4, 2) + 0.1, mgp = c(3, 1, 0))

hist(t_permuted_res, "FD", freq = FALSE, main = NA, 
     xlab = "Permutation t Test Statistics (Wind)", ylab = "Density",
     col = "lightgray", border = "white")

curve(dt(x, df = df_t), add = TRUE, lty = 2, lwd = 2, col = 1)
abline(v = c(qt(alpha/2, df = df_t), qt(1 - alpha/2, df = df_t)), col = 2, lty = 2, lwd = 4)
abline(v = c(quantile(t_permuted_res, alpha/2), quantile(t_permuted_res, 1 - alpha/2)), col = 4, lty = 2, lwd = 4)
abline(v = c(t_observed, -t_observed), col = 3, lty = 2, lwd = 4)
legend("topright", 
       c("Theoretical Distribution", "Theoretical Quantile", 
         "Empirical Quantile", "Observed Statistic"),
       col = c(1, 2, 4, 3), lty = rep(2, 4), lwd = rep(4, 4), 
       cex = 0.8, bg = "white")
mtext(text = paste("Observed t (Wind):", round(t_observed, 2), 
                   "   Estimated P-value:", round(p_val_t_wind_precise, 5)), 
      side = 1, line = -1.5, adj = 0.5, outer = FALSE, cex = 0.9, col = "black")
```

The estimated p values are larger than the theoretical p values under null hypothesis, thus the model, as well as the covariates `wind` seem to be not statistically significant.

## Transformation

Transformation is vital for our data.

```{r}
library(MASS)

df_quant_clean <- subset(df, select = -c(month, day))

bc_model <- lm(area ~ X + Y + FFMC + DMC + DC + ISI + temp + RH + wind,
               data = df_quant_clean)

bc <- boxcox(bc_model,
             lambda = seq(-2, 2, by = 0.01),
             plotit = TRUE)

best_lambda <- bc$x[which.max(bc$y)]
print(best_lambda)
```

Thus, we can use the log transformation to `area` .

Because the area in this data set is strictly larger than 0, `log(area)` is enough and make more sense in this situation. We also test the r-squared and adjusted r-squared, though both are small, but log transform is better.

```{r}
df_quant$trans_area = log(df_quant$area)
hist(df_quant$trans_area)
```

```{r}
ols_trans <- lm(trans_area ~ . - area, data = df_quant)
summary(ols_trans)
plot(ols_trans, which = 1)
```

```{r}
df_quant$trans_area = log(df_quant$area)
ols_log = lm(trans_area ~ . - area, data = df_quant)

df_quant$trans_area = log1p(df_quant$area)
ols_log1p = lm(trans_area ~ . - area, data = df_quant)

compare = data.frame(
  model = c("log", "log1p"),
  n = c(nobs(ols_log), nobs(ols_log1p)),
  p = c(length(coef(ols_log)) - 1, length(coef(ols_log1p)) - 1),
  r2 = c(summary(ols_log)$r.squared, summary(ols_log1p)$r.squared),
  adj_r2 = c(summary(ols_log)$adj.r.squared, summary(ols_log1p)$adj.r.squared)
)

compare
```

### Transformation of Covariates

The fire weather variables `DMC`, `DC`, and `ISI` are positive indexes and can be highly right-skewed. The wind speed is also positive. Therefore, we use `log1p()` for these covariates, because it keeps the transformation stable even if a future data set contains zeros. For `FFMC`, most observations are close to the upper bound 100, so we use the log distance from the upper bound.

```{r}
df$month <- factor(df$month,
                   levels = c("jan", "feb", "mar", "apr", "may", "jun",
                              "jul", "aug", "sep", "oct", "nov", "dec"),
                   ordered = TRUE)

df$day <- factor(df$day,
                 levels = c("mon", "tue", "wed", "thu", "fri", "sat", "sun"),
                 ordered = TRUE)

num_vars <- names(df)[sapply(df, is.numeric)]
cat_vars <- names(df)[!sapply(df, is.numeric)]
```

```{r}
par(mfrow = c(3, 4), mar = c(4, 4, 3, 1))

for (v in num_vars) {
  hist(df[[v]],
       main = v,
       xlab = v,
       col = "lightgray",
       border = "black")
}
```

```{r}
par(mfrow = c(1, length(cat_vars)), mar = c(6, 4, 3, 1))

for (v in cat_vars) {
  barplot(table(df[[v]]),
          main = v,
          ylab = "Frequency",
          col = "lightgray",
          border = "black",
          las = 2)
}
```

From the plots, we think transformation for covariates might not be meaningful.

```{r}
library(qqconf)

r = rstandard(ols_trans)

qq_conf_plot(r, points_params = list(pch = 16, cex = 0.5))
```

Since all standardized residuals appear to fall within the 95% simultaneous confidence bands, we fail to reject the global null hypothesis of normality. Thus, the standardized residuals from the transformed linear model are reasonably consistent with a standard normal distribution.


### Interaction Effects

```{r}
qvar = "month"

df_trans[[qvar]] = factor(df_trans[[qvar]])
df_trans[[qvar]] = droplevels(df_trans[[qvar]])

additive_formula = as.formula(
  paste("log_area ~ X + Y + FFMC + DMC + DC + ISI + temp + RH + wind +", qvar)
)

interaction_formula = as.formula(
  paste("log_area ~ (X + Y + FFMC + DMC + DC + ISI + temp + RH + wind) *", qvar)
)

additive_model = lm(additive_formula, data = df_trans)
interaction_model = lm(interaction_formula, data = df_trans)

anova(additive_model, interaction_model)
summary(interaction_model)
```

The full interaction model using month creates a large number of interaction terms. Several coefficients are not estimable because of singularities, suggesting that the month-level interaction model is overparameterized. Therefore, we combine months into broader fire-season categories and investigate interaction effects using this simplified qualitative covariate.

```{r}
df_trans$fire_season = ifelse(df_trans$month %in% c("jul", "aug", "sep"),
                              "Peak",
                              "OffPeak")

df_trans$fire_season = factor(df_trans$fire_season)
table(df_trans$fire_season)
```
```{r}
additive_model = lm(log_area ~ X + Y + FFMC + DMC + DC + ISI + temp + RH + wind + fire_season,
                    data = df_trans)
summary(additive_model)

interaction_model = lm(log_area ~ (X + Y + FFMC + DMC + DC + ISI + temp + RH + wind) * fire_season,
                       data = df_trans)
summary(interaction_model)

interaction_test = anova(additive_model, interaction_model)
print(interaction_test)
```