---
title: "Fire"
author: "Xinlu Jiang"
date: "2026-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(car);library(lmtest);library(sandwich);library(sandwich);library(lmtest)
```

## Introduction to "Fire"

This data set contains measurements on 270 forest fires in the northeast region of Portugal. We first read in the data and check the dimension, variable names, first few rows, and summary statistics. 

```{r}
## Introduction to Fire Data

df = read.table("fire.txt", header = TRUE)

dim(df)
head(df)
summary(df)
```

```{r}
names(df)
```
Our response variable is `area`, by our first eye, we noticed that it has a median of 6.37, a mean of 24.6, while the 3rd qu of 15.42, indicating that it does not follow normal distribution for sure and we might transform it into log form in the future. Quantitative covariates includes `FFMC`, `DMC`, `DC`, `ISI`, `temp`, `RH`, `wind`, and location information `X` and `Y`. While the rest, `month` and `day` are qualitative variables that will be used later for fixed/random effects.

```{r}
df$month = factor(df$month)
df$day = factor(df$day)

summary(df)
```
The variables month and day are qualitative variables, so I convert them into factors. I do not include them in the first baseline model because we want tto use all quantitative covariates and ignore qualitative variables.

```{r}
df_quant = df[, c("X", "Y", "FFMC", "DMC", "DC", "ISI",
                  "temp", "RH", "wind", "area")]

summary(df_quant)
```
The quantitative covariates are X, Y, FFMC, DMC, DC, ISI, temp, RH, and wind. The response variable is area. I keep only these variables in df_quant.

```{r}
Y = df_quant$area

hist(Y, xlab = "Area", main = "Histogram of Area")
boxplot(Y, ylab = "Area", pch = 16)
```
The response variable area is strongly right-skewed. The mean is much larger than the median, and the histogram shows that most fires have small burned area while a few fires have much larger values. This means the normality assumption may be questionable, so a transformation may be useful later. For now, I still start with the required baseline linear model using the original response.
```{r}
plot(df_quant, pch = 16, cex = 0.5)
```
The scatterplot matrix gives an overall view of the relationships between all quantitative variables. From the plot, most variables do not show strong linear relationships with the response variable area. The values of area are heavily concentrated near zero, with a few large values, which confirms that the response variable is highly right-skewed.

Some covariates, such as DMC, DC, and ISI, show moderate positive relationships with each other, indicating possible correlation among predictors. The variables temp and RH appear to have a negative relationship, which is also visible in the scatterplots.

Overall, the relationships between predictors and the response are weak and noisy, which suggests that the linear model may have low explanatory power.


## Basic Model

### OLS and tests

```{r}
ols_model = lm(area ~ ., df_quant)
summary(ols_model)
```

we fit a multiple linear regression model using area as the response and all quantitative covariates as predictors.

From the coefficient table, none of the covariates are statistically significant at the 5% level, since all p-values are greater than 0.05. The signs of the coefficients suggest weak relationships, but they are not statistically reliable.

The overall F-statistic has a p-value of 0.4355, which means we fail to reject the null hypothesis that all regression coefficients are equal to zero. This indicates that the model does not provide strong evidence that the predictors explain the variation in area.

The multiple R^2 is about 0.0337, and the adjusted R^2 is extremely small (approximately 0.0002), which shows that the model explains almost none of the variability in the response.

Overall, this suggests that the linear relationship between the covariates and the response is very weak, which is consistent with the earlier scatterplot observations.


```{r}
Y = df_quant$area
X = model.matrix(ols_model)

n = dim(X)[1]
p = dim(X)[2]

colnames(X)
```
I define the response vector Y, the design matrix X, the sample size n, and the number of estimated regression coefficients p. The first column of X is the intercept column. I print colnames(X) because the order of the columns matters when I build the constraint matrix R.

```{r}
beta = solve(crossprod(X), crossprod(X, Y))[,]
print(beta)

beta = ols_model$coefficients
print(beta)

residuals = Y - X %*% beta

sigma = sqrt(mean(residuals^2))
print(sigma)

S = sqrt(sum(residuals^2) / (n - p))
print(S)

S = summary(ols_model)$sigma
print(S)
```
I compute the OLS coefficients using the matrix formula and then compare them with ols_model$coefficients. They should match. I also compute the MLE estimate of residual standard deviation, sigma, and the unbiased residual standard error, S. The later exact F and Wald tests use this same setup.

Our required linear constrain to test is 
$$
H_0:
\begin{aligned}
50\beta_{DMC} + 10\beta_{RH} &= 3 \\
6\beta_{ISI} + 12\beta_{temp} &= 3
\end{aligned}
$$

```{r}
## Linear Constraints

R = rbind(c(0, 0, 0, 0, 50, 0, 0, 0, 10, 0),
          c(0, 0, 0, 0, 0, 0, 6, 12, 0, 0))

r = c(3, 3)
k = length(r)

print(colnames(X))
print(R)
print(r)
```

```{r}
## Exact F Test

FT = crossprod(R %*% beta - r,
               solve(R %*% solve(crossprod(X), t(R)),
                     R %*% beta - r))[,] / (k * S^2)

print(FT)

pvalue_F = pf(FT, k, n - p, lower.tail = FALSE)
print(pvalue_F)

## Wald Test

WT = k * FT
print(WT)

pvalue_WT = pchisq(WT, k, lower.tail = FALSE)
print(pvalue_WT)
```

The results from the exact F test and the asymptotic Wald test are very similar. The p-values are both very large (close to 1), so we fail to reject the null hypothesis at the 5% significance level.

This means there is no statistical evidence that the given linear constraints are violated, so the null hypothesis is consistent with the data.

The agreement between the F test and the Wald test is expected, since both tests are based on the same null hypothesis and use the same estimated coefficients. The Wald test is an asymptotic approximation of the exact F test, so when the sample size is large, the two tests tend to give similar conclusions.

### Heteroscedasticity

We need homoscedasticity to calculate a valid OLS SE, so we can plot a scatter plot to watch it, or use BP test as well. We found out a easier way to draw the residual plots by <https://rpubs.com/sharmaar3/Residual_Plots>.

```{r}
plot(ols_model, which = 1)
```

From the plot, most residuals are centered around 0 with no clear funnel shape.There are a few large outliers, but overall variance looks roughly constant.So visually, there is no strong evidence of heteroscedasticity.

```{r}
# Step 1: residuals
residuals = ols_model$residuals

# Step 2: squared residuals
Y_aux = residuals^2

# Step 3: regress squared residuals on covariates
X = model.matrix(ols_model)
aux_model = lm(Y_aux ~ X[,-1])

# Step 4: compute R^2
R2_aux = summary(aux_model)$r.squared

# Step 5: compute BP statistic
n = length(residuals)
BP_stat = n * R2_aux

# Step 6: p-value
p_value = 1 - pchisq(BP_stat, df = ncol(X) - 1)

print(BP_stat)
print(p_value)
```

We formally test for heteroscedasticity using the Breusch–Pagan test. This test is implemented by regressing the squared residuals on the covariates and computing the test statistic nR^2, which follows a chi-squared distribution under the null hypothesis of constant variance .

From our computation, we obtain:

BP=5.399977
p-value = 0.7981412

Since the p-value is much larger than 0.05, we fail to reject the null hypothesis. Therefore, there is no statistical evidence that the variance of the residuals depends on the covariates.

This result is consistent with the residual plot, where the spread of residuals appears relatively stable across fitted values, despite the presence of a few large outliers.

However, we doubt the result of both tests, because we think most of our respond data are around 0, it is a serious right skewed data. And because BP test rely heavily on normality assumption, it may not work in our basic model. 

## Robust Standard Errors 

```{r}
vcovOLS = summary(ols_model)$sigma^2 * summary(ols_model)$cov.unscaled
print(vcovOLS)

Sigma = diag(residuals^2 / (1 - hatvalues(ols_model))^2)

vcovHC3 = summary(ols_model)$cov.unscaled %*% crossprod(X, Sigma) %*%
  X %*% summary(ols_model)$cov.unscaled

print(vcovHC3)

```


We compare the usual OLS covariance matrix with the HC3 version to see if accounting for heteroscedasticity changes anything. From the results, the values in the two matrices are very similar overall, with only small differences across some coefficients. This tells us that adjusting for heteroscedasticity does not significantly change how much uncertainty we estimate for the coefficients. In other words, the variability of the estimates is pretty stable. This is consistent with what we saw earlier from the residual plot and the Breusch–Pagan test, where there was no strong evidence of heteroscedasticity. So overall, using the HC3 adjustment does not change our conclusions, and the OLS results appear to be reliable.

```{r}

SE = cbind(
  OLS = sqrt(diag(vcovOLS)),
  HC3 = sqrt(diag(vcovHC3))
)

print(SE)
```
From the comparison, we notice that many of the HC3 standard errors are actually smaller than the OLS standard errors, which might seem unexpected at first. This can happen when there is little to no strong heteroscedasticity in the data, since the HC3 adjustment is designed to correct for uneven variance but does not always inflate the standard errors. In some cases, the adjustment can slightly reduce the estimated variability, especially when certain observations have high leverage and the correction redistributes their influence. Overall, since the differences are not large and the pattern is fairly consistent across variables, this suggests that heteroscedasticity is not a major issue in this model. This again supports our earlier findings that the OLS results are relatively stable and reliable.

```{r}
alpha = 0.05

CI_HC3 = cbind(
  beta - qt(1 - alpha / 2, n - p) * sqrt(diag(vcovHC3)),
  beta + qt(1 - alpha / 2, n - p) * sqrt(diag(vcovHC3))
)

colnames(CI_HC3) = colnames(confint(ols_model))
print(CI_HC3)

CI_OLS = confint(ols_model)
print(CI_OLS)
```

Looking at the 95% confidence intervals, the HC3 intervals and the OLS intervals are very similar overall. For most variables, the HC3 intervals are slightly narrower, which matches what we saw earlier with the standard errors being a bit smaller. More importantly, all the intervals still include 0, so the conclusion about which variables are statistically significant does not change at all. This tells us that adjusting for heteroscedasticity does not really affect our inference. Overall, the results are very consistent between the two methods, which again suggests that heteroscedasticity is not a major issue in this model.

```{r}
test = matrix(0, p, 4)
rownames(test) = names(ols_model$coefficients)
colnames(test) = colnames(summary(ols_model)$coefficients)

test[,1] = beta
test[,2] = sqrt(diag(vcovHC3))
test[,3] = test[,1] / test[,2]
test[,4] = 2 * pt(abs(test[,3]), n - p, lower.tail = FALSE)

print(test)

summary(ols_model)
```

Looking at the robust t-test results compared to the usual OLS output, the overall conclusions stay the same. The standard errors change a bit (mostly getting smaller), which makes some t-values slightly larger, but not enough to make any variables statistically significant. All the p-values are still well above 0.05, so none of the covariates show a meaningful relationship with the response variable under either method. This tells us that adjusting for heteroscedasticity does not change our inference at all, and the results from the model are consistent and stable.

### Iteratively Reweighted Least Squares

```{r}
plot(rstandard(ols_model) ~ temp, df_quant,
     ylab = "Standardized Residuals",
     pch = 16)
abline(h = 0, lty = 2, lwd = 2)
```

```{r}
# IRLS with residual variance modeled as a function of temp

irls_model = ols_model
residuals_irls = irls_model$residuals

w = rep(1, nrow(df_quant))
err = Inf
tol = 1e-10
niter = 0
maxiter = 1e3

while (err > tol & niter < maxiter)
{
  
  aux = lm(log(residuals_irls^2) ~ log(temp), df_quant)
  err = sum(abs(w - exp(-aux$fitted.values))) / sum(w)
  w = exp(-aux$fitted.values)
  WLS = lm(area ~ ., df_quant, weights = w)
  residuals_irls = WLS$residuals
  niter = niter + 1
  
}

summary(aux)
summary(WLS)
```

```{r}
gamma_hat = coef(aux_model)

alpha_0_hat = exp(gamma_hat[1])
alpha_1_hat = gamma_hat[2]

alpha_0_hat
alpha_1_hat

cat(
  paste0(
    "Estimated residual variance equation: sigma_i^2 = ",
    round(alpha_0_hat, 4),
    " * temp_i^",
    round(alpha_1_hat, 4)
  )
)
```

```{r}
ols_summary = summary(ols_model)
irls_summary = summary(irls_model)

compare_ols_irls = data.frame(
  Variable = names(coef(ols_model)),
  OLS_Estimate = coef(ols_model),
  OLS_SE = ols_summary$coefficients[, "Std. Error"],
  OLS_t = ols_summary$coefficients[, "t value"],
  OLS_p_value = ols_summary$coefficients[, "Pr(>|t|)"],
  IRLS_Estimate = coef(irls_model),
  IRLS_SE = irls_summary$coefficients[, "Std. Error"],
  IRLS_t = irls_summary$coefficients[, "t value"],
  IRLS_p_value = irls_summary$coefficients[, "Pr(>|t|)"]
)

print(compare_ols_irls)
```

```{r}
ols_ci = confint(ols_model, level = 0.95)
irls_ci = confint(irls_model, level = 0.95)

print("--- OLS 95% CI ---")
print(ols_ci)

print("--- IRLS/WLS 95% CI ---")
print(irls_ci)
```

```{r}
summary(w)
```


```{r}
plot(WLS$fitted.values, rstandard(WLS),
     xlab = "Fitted Values",
     ylab = "Standardized Residuals",
     pch = 16)
abline(h = 0, lty = 2, lwd = 2)
```
Compared with OLS, the IRLS estimates change substantially for several coefficients.
 For example, the coefficient of `temp` decreases from 1.1524 under OLS to 0.1311 
 under IRLS, and the coefficient of `wind` changes from 2.5191 to -0.1492. The standard
 errors are also generally smaller under IRLS. However, the inferential conclusion remains
the same: none of the covariates is statistically significant at the 5% level.

The confidence intervals from IRLS are generally narrower than the OLS confidence intervals.
 For example, the 95% CI for `temp` changes from \([-1.6071, 3.9118]\) under OLS to 
 \([-0.9781, 1.2402]\) under IRLS. Similarly, the 95% CI for `wind` changes from 
 \([-3.6896, 8.7279]\) under OLS to \([-2.5885, 2.2900]\) under IRLS. Even though 
 the IRLS intervals are narrower, they still contain zero for every covariate.

## Bootstrap and Jackknife 

```{r}
## Jackknife

n = nrow(df_quant)
p = length(coef(ols_model))

beta_jack = matrix(0, n, p)

for (i in 1:n)
{
  fit_i = lm(area ~ ., df_quant[-i, ])
  beta_jack[i, ] = coef(fit_i)
}

beta_bar = colMeans(beta_jack)

se_jack = sqrt((n - 1) / n *
  colSums((sweep(beta_jack, 2, beta_bar, "-"))^2))

print(se_jack)

## Bootstrap

set.seed(0)
B = 1000

beta_boot = matrix(0, B, p)

for (b in 1:B)
{
  idx = sample(1:n, n, replace = TRUE)
  fit_b = lm(area ~ ., df_quant[idx, ])
  beta_boot[b, ] = coef(fit_b)
}

se_boot = apply(beta_boot, 2, sd)

print(se_boot)
```

```{r}

se_ols = summary(ols_model)$coefficients[,2]
alpha = 0.05
beta_hat = coef(ols_model)


CI_normal = cbind(
  beta_hat - qnorm(1 - alpha/2) * se_boot,
  beta_hat + qnorm(1 - alpha/2) * se_boot
)


CI_percentile = apply(beta_boot, 2, quantile, probs = c(alpha/2, 1 - alpha/2))
CI_percentile = t(CI_percentile)


CI_pivot = cbind(
  2 * beta_hat - CI_percentile[,2],
  2 * beta_hat - CI_percentile[,1]
)


var_names = names(beta_hat)

final_table = data.frame(
  Variable = var_names,
  OLS_SE = se_ols,
  Jackknife_SE = se_jack,
  Bootstrap_SE = se_boot,
  Normal_L = CI_normal[,1],
  Normal_U = CI_normal[,2],
  Percentile_L = CI_percentile[,1],
  Percentile_U = CI_percentile[,2],
  Pivot_L = CI_pivot[,1],
  Pivot_U = CI_pivot[,2]
)


final_table[,-1] = round(final_table[,-1], 3)

print(final_table)
```
Overall, the resampling results show that the bootstrap and jackknife standard errors are quite close to each other and generally similar to the OLS standard errors. This suggests that the coefficient estimates are fairly stable and not overly sensitive to the specific sample. In most cases, the bootstrap and jackknife SEs are slightly smaller than the OLS SEs, which can happen when the model assumptions are not strongly violated or when the data does not exhibit strong heterogeneity.

Looking at the confidence intervals, all three methods (normal, percentile, and pivotal) give very similar ranges for most variables. The percentile and pivotal intervals are almost identical in many cases, which indicates that the sampling distribution of the estimates is fairly symmetric. Compared to the normal intervals, the bootstrap-based intervals are sometimes slightly shifted or adjusted, reflecting small deviations from normality in the data.

Importantly, across all methods, most confidence intervals include zero, meaning that none of the covariates appear to be statistically significant. This is consistent with the earlier OLS results and supports the conclusion that the model has very weak explanatory power for this dataset.

In terms of purpose, the jackknife and bootstrap are used to check the reliability of the standard errors and confidence intervals without relying heavily on model assumptions. Here, since all methods lead to very similar conclusions, it reinforces that the lack of significance is not due to estimation method issues, but rather because the underlying relationships in the data are weak.



### Permutation test of model and `Wind`
```{r}
### Permutation test of full model and wind

set.seed(0)
alpha = 0.05
B = 10000

Y = df_quant$area
X = model.matrix(ols_model)
n = dim(X)[1]
p = dim(X)[2] - 1
```

## Permutation F test: overall model significance
```{r}
## Permutation F test: overall model significance

f_observed = summary(ols_model)$fstatistic[1]
df_f1 = summary(ols_model)$fstatistic[2]
df_f2 = summary(ols_model)$fstatistic[3]

p_val_F_theory = pf(f_observed, df_f1, df_f2, lower.tail = FALSE)

f_permuted = numeric(B)

for (i in 1:B)
{
  perm_y = sample(Y)
  perm_model = lm(perm_y ~ ., data = df_quant[, names(df_quant) != "area"])
  f_permuted[i] = summary(perm_model)$fstatistic[1]
}

p_val_F_perm = (1 + sum(f_permuted >= f_observed)) / (1 + B)

print(f_observed)
print(p_val_F_theory)
print(p_val_F_perm)
```

```{r}
## Plot for permutation F test

hist(f_permuted, "FD", freq = FALSE, main = NA,
     xlab = "Permutation F Test Statistics",
     ylab = "Density")

curve(df(x, df_f1, df_f2), add = TRUE, lty = 2, lwd = 4)

abline(v = qf(1 - alpha, df_f1, df_f2), col = 2, lty = 2, lwd = 4)
abline(v = quantile(f_permuted, 1 - alpha), col = 4, lty = 2, lwd = 4)
abline(v = f_observed, col = 7, lty = 2, lwd = 4)

legend("topright",
       c("Theoretical Distribution", "Theoretical Quantile",
         "Empirical Quantile", "Observed Statistic"),
       col = c(1, 2, 4, 7),
       lty = rep(2, 4),
       lwd = rep(4, 4))
```
The permutation F test is used to check whether the model is statistically significant overall, meaning whether all the covariates together have any real effect on the response variable. The observed F statistic is about 1.006, which is very small. The theoretical p-value is around 0.436, and the permutation-based p-value is about 0.395. These two values are pretty close to each other, which means the permutation test is giving a similar conclusion as the usual F test.

Looking at the plot, the observed F statistic (yellow line) falls well inside the bulk of the permutation distribution, not in the extreme tail. This visually confirms that the observed value is not unusual under the null hypothesis. Also, both the theoretical and empirical critical values are clearly to the right of the observed statistic, which further shows that we do not have enough evidence to reject the null.

Since both p-values are much larger than 0.05, we fail to reject the null hypothesis. This means the model is not statistically significant overall, and the covariates do not jointly explain the variation in the response variable. The permutation result supports the same conclusion as the theoretical test, so there is no evidence that the model has meaningful predictive power in this dataset.

Permutaion t test of covariate "wind":

We chose wind to do the permutaion t test because, in this fire dataset, wind is one of the most physically relevant variables for fire spread, so it’s meaningful to test it individually. At the same time, our earlier OLS results show that wind is not statistically significant (small t-value, large p-value), making it a good candidate to verify whether this lack of significance still holds under a nonparametric permutation test. This helps confirm that the conclusion is not just due to model assumptions, but is consistent with the data itself.
```{r}
## Permutation t test for wind

model_reduced = lm(area ~ . - wind, data = df_quant)

fitted_reduced = model_reduced$fitted.values
res_reduced = model_reduced$residuals

t_observed = summary(ols_model)$coefficients["wind", "t value"]
df_t = summary(ols_model)$df[2]

p_val_t_theory = 2 * pt(abs(t_observed), df_t, lower.tail = FALSE)

t_permuted = numeric(B)

for (i in 1:B)
{
  res_star = sample(res_reduced)
  y_star = fitted_reduced + res_star
  
  temp_data = df_quant
  temp_data$area = y_star
  
  model_star = lm(area ~ ., data = temp_data)
  t_permuted[i] = summary(model_star)$coefficients["wind", "t value"]
}

p_val_t_perm = (1 + sum(abs(t_permuted) >= abs(t_observed))) / (1 + B)

print(t_observed)
print(p_val_t_theory)
print(p_val_t_perm)
```

```{r}
## Plot for permutation t test

hist(t_permuted, "FD", freq = FALSE, main = NA,
     xlab = "Permutation t Test Statistics (Wind)",
     ylab = "Density")

curve(dt(x, df_t), add = TRUE, lty = 2, lwd = 4)

abline(v = qt(c(alpha / 2, 1 - alpha / 2), df_t),
       col = 2, lty = 2, lwd = 4)

abline(v = quantile(t_permuted, c(alpha / 2, 1 - alpha / 2)),
       col = 4, lty = 2, lwd = 4)

abline(v = c(t_observed, -t_observed),
       col = 7, lty = 2, lwd = 4)

legend("topright",
       c("Theoretical Distribution", "Theoretical Quantile",
         "Empirical Quantile", "Observed Statistic"),
       col = c(1, 2, 4, 7),
       lty = rep(2, 4),
       lwd = rep(4, 4))
```
The permutation t test is used to check whether the single covariate wind has a statistically significant effect on the response variable. The observed t statistic is about 0.799, which is very close to 0, indicating a weak effect. The theoretical p-value is about 0.425, and the permutation-based p-value is about 0.415. These two values are very similar, which shows that the permutation test agrees with the classical t-test result.

From the plot, the observed t statistic (yellow line) is located near the center of the permutation distribution, not in the tails. This means the observed value is very typical under the null hypothesis. The empirical quantiles (blue) and theoretical quantiles (red) are also very close, which suggests that the theoretical t distribution is a good approximation in this case.

Since both p-values are much larger than 0.05, we fail to reject the null hypothesis. This means that the covariate wind is not statistically significant in explaining the response variable. The permutation test confirms the same conclusion as the theoretical test, so there is no evidence that wind has a meaningful effect in this dataset.


```{r}
## Summary table

perm_table = data.frame(
  Test = c("Overall F test", "wind t test"),
  Observed_Statistic = c(f_observed, t_observed),
  Theoretical_P_Value = c(p_val_F_theory, p_val_t_theory),
  Permutation_P_Value = c(p_val_F_perm, p_val_t_perm)
)

print(perm_table)
```


Both the overall F test and the wind t test show that the permutation p-values are very close to the theoretical p-values, and all of them are much larger than 0.05. This means we fail to reject the null hypothesis in both cases.

For the overall model, this suggests that the set of quantitative covariates does not provide strong explanatory power for the response variable. For the wind covariate, the result confirms that its effect is not statistically significant.

Since the permutation results agree closely with the theoretical results, this indicates that our conclusion is robust and not driven by distributional assumptions. In the context of this dataset, it suggests that the weak relationships we observed (very low signal) are real, rather than artifacts of model assumptions.


## Transformation

Transformation is vital for our data.

```{r}
library(MASS)

df_quant_clean <- subset(df, select = -c(month, day))

bc_model <- lm(area ~ X + Y + FFMC + DMC + DC + ISI + temp + RH + wind,
               data = df_quant_clean)

bc <- boxcox(bc_model,
             lambda = seq(-2, 2, by = 0.01),
             plotit = TRUE)

best_lambda <- bc$x[which.max(bc$y)]
print(best_lambda)
```
Thus, we can use the log transformation to `area` .

Because the area in this data set is strictly larger than 0, `log(area)` is enough and make more sense in this situation. We also test the r-squared and adjusted r-squared, though both are small, but log transform is better.

By applying the log transformation, we reduce the impact of large outliers and compress the scale of extreme values. This helps make the distribution more symmetric and improves the behavior of residuals. Overall, the transformation makes the linear model assumptions more reasonable and leads to more reliable inference.

```{r}
df_quant$trans_area = log(df_quant$area)
hist(df_quant$trans_area)
```
After applying the log transformation to area, the histogram shows a much more symmetric distribution compared to the original data. The values are now more centered and less dominated by extreme large observations, which reduces the strong right skew we saw before.

Most of the data is concentrated in the middle range, and the tails are less stretched out. This indicates that the transformation successfully compresses large values and makes the distribution closer to normal.

Overall, this improvement in the shape of the response variable suggests that the log transformation helps make the data better suited for linear regression assumptions, especially normality and constant variance.


```{r}
ols_trans <- lm(trans_area ~ . - area, data = df_quant)
summary(ols_trans)
plot(ols_trans, which = 1)
```
From the residuals vs fitted plot after applying the log transformation, the residuals are more randomly scattered around zero with no strong pattern. The red smooth line is relatively flat, which suggests that the linearity assumption is reasonably satisfied.

Compared to the original model, there is less visible structure in the residuals, meaning the model fit has improved. The spread of residuals also looks more consistent across fitted values, indicating that the variance is more stable.

Although a few points still appear as mild outliers, the overall pattern does not show clear heteroscedasticity or curvature. This suggests that the log transformation helps improve the model assumptions and provides a better fit for the data.

```{r}
df_quant$trans_area = log(df_quant$area)
ols_log = lm(trans_area ~ . - area, data = df_quant)

df_quant$trans_area = log1p(df_quant$area)
ols_log1p = lm(trans_area ~ . - area, data = df_quant)

compare = data.frame(
  model = c("log", "log1p"),
  n = c(nobs(ols_log), nobs(ols_log1p)),
  p = c(length(coef(ols_log)) - 1, length(coef(ols_log1p)) - 1),
  r2 = c(summary(ols_log)$r.squared, summary(ols_log1p)$r.squared),
  adj_r2 = c(summary(ols_log)$adj.r.squared, summary(ols_log1p)$adj.r.squared)
)

compare
```
From the comparison table, both the log and log1p transformations produce very small R^2 and adjusted R^2, indicating that the overall model still has weak explanatory power. However, the log transformation performs slightly better than log1p in both metrics.

Specifically, the log model has a higher R^2 and a noticeably higher adjusted R^2, while the log1p model shows a drop in adjusted R^2, suggesting a worse fit after accounting for model complexity.

Since both models use the same number of predictors, the difference in adjusted R^2 directly reflects model performance rather than overfitting. Therefore, the log transformation is preferred over log1p for this dataset.

Overall, although the improvement is small, the log transformation provides a slightly better fit and is more appropriate for modeling the response variable.

```{r}

r <- rstandard(ols_trans)

qqnorm(r,
       main = "Normal Q-Q Plot of Standardized Residuals",
       pch = 16)

qqline(r,
       col = "red",
       lwd = 2)
```
From the Q-Q plot of standardized residuals, most of the points lie close to the reference line, especially in the middle range. This suggests that the residuals are approximately normally distributed.

There are some slight deviations at the tails, where a few points move away from the line, indicating mild departures from normality for extreme values. However, these deviations are not very large.

Overall, the plot shows that the normality assumption is reasonably satisfied after the transformation, and the residuals are fairly consistent with a normal distribution.

### Transformation of Covariates

The fire weather variables `DMC`, `DC`, and `ISI` are positive indexes and can be highly right-skewed. The wind speed is also positive. Therefore, we use `log1p()` for these covariates, because it keeps the transformation stable even if a future data set contains zeros. For `FFMC`, most observations are close to the upper bound 100, so we use the log distance from the upper bound.

```{r}
df$month <- factor(df$month,
                   levels = c("jan", "feb", "mar", "apr", "may", "jun",
                              "jul", "aug", "sep", "oct", "nov", "dec"),
                   ordered = TRUE)

df$day <- factor(df$day,
                 levels = c("mon", "tue", "wed", "thu", "fri", "sat", "sun"),
                 ordered = TRUE)

num_vars <- names(df)[sapply(df, is.numeric)]
cat_vars <- names(df)[!sapply(df, is.numeric)]
```

```{r}
par(mfrow = c(3, 4), mar = c(4, 4, 3, 1))

for (v in num_vars) {
  hist(df[[v]],
       main = v,
       xlab = v,
       col = "lightgray",
       border = "black")
}
```

```{r}
par(mfrow = c(1, length(cat_vars)), mar = c(6, 4, 3, 1))

for (v in cat_vars) {
  barplot(table(df[[v]]),
          main = v,
          ylab = "Frequency",
          col = "lightgray",
          border = "black",
          las = 2)
}
```

From the plots, we think transformation for covariates might not be meaningful.

## interaction 
```{r}
df = read.table("fire.txt", header = TRUE)

df_trans = df

df_trans$month = factor(df_trans$month)
df_trans$day = factor(df_trans$day)
df_trans$log_area = log(df_trans$area)

```

```{r}
df_inter = df_trans[, c(
  "log_area", "day", "month",
  "X", "Y", "FFMC", "DMC", "DC", "ISI", "temp", "RH", "wind"
)]

summary(df_inter)
```

First, we test the interaction of day with other covariates, expect month. We have 200 degrees of freedom in this model.

```{r}
fit_day_inter = lm(
  log_area ~ day * (X + Y + FFMC + DMC + DC + ISI + temp + RH + wind),
  data = df_inter
)

summary(fit_day_inter)
anova(fit_day_inter)
```

Although most variables and interactions are not statistically significant, the interaction between day and DMC shows some evidence of significance. This suggests that the effect of fuel moisture on fire area may vary across days. However, the overall model fit remains weak, indicating that interaction terms alone are not sufficient to explain the variability in the data.

```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:FFMC)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:temp)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - temp)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:RH)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - FFMC)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:DC)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:wind)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - DC)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:Y)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - Y)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:X)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - X)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - day:ISI)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - RH)
anova(fit_day_inter)
```
```{r}
fit_day_inter = update(fit_day_inter, . ~ . - wind)
anova(fit_day_inter)
```
We start with a full interaction model between day and all the quantitative variables, then go step by step removing terms that aren’t significant based on the ANOVA table. Most of the interaction terms (like day:X, day:Y, day:FFMC, day:temp, day:RH, day:wind, and day:DC) all have pretty large p-values and small F values, so they’re not really contributing much to the model. When we remove them, the residual mean square stays around the same (~2.2), which shows that dropping them doesn’t hurt the model fit. The only interaction that keeps showing up as significant is day:DMC, with p-values around 0.02–0.04, so this suggests that the effect of DMC on fire size actually depends on the day. Also, DMC by itself isn’t significant, which means its effect is not constant and only shows up through the interaction. ISI is kind of borderline significant (around 0.06), so we might keep it depending on how strict we want to be. In the end, we simplify the model a lot and keep the important parts, which are day, DMC, maybe ISI, and especially the interaction day:DMC, since that’s the only clear pattern we see from the results.
We also test interaction effects of month, yet nothing is significant.

## Mixed effect model 

```{r}
library(lme4)

df_mixed = df

df_mixed$month = factor(df_mixed$month)
df_mixed$day = factor(df_mixed$day)


df_mixed$log_area = log(df_mixed$area)


fit_mixed = lmer(
  log_area ~ X + Y + FFMC + DMC + DC + ISI + temp + RH + wind +
    (1 | day) + (1 | month),
  data = df_mixed
)

summary(fit_mixed, correlation = FALSE)


```
Overall, from the mixed-effects model, we can see that most of the fixed effects are not statistically strong, since their t-values are pretty small and close to zero. The only variable that shows some signal is DMC, which has a relatively larger t-value compared to the others, but even that is not very strong. This matches what we saw earlier with the OLS models, where it was already hard to find meaningful predictors for fire area. For the random effects, the variance for day is basically zero, while month has a very small variance, meaning these grouping variables do not explain much of the variation in the response. Most of the variation is still coming from the residual, which suggests that the model is not capturing the main drivers of fire size. Also, the “boundary (singular) fit” warning tells us that one of the random effects (in this case day) is not really contributing to the model, which makes sense since its variance is almost zero. Overall, this confirms that even after using a mixed-effects model, the data is still very noisy and difficult to explain, and the main patterns in fire area are likely driven by factors not included in the dataset.
```{r}

full = lmer(
  log_area ~ X + Y + FFMC + DMC + DC + ISI + temp + RH + wind +
    (1 | day) + (1 | month),
  data = df_mixed,
  REML = FALSE
)

fixed_names = c("X", "Y", "FFMC", "DMC", "DC", "ISI", "temp", "RH", "wind")

LR = numeric(length(fixed_names))
pvalue = numeric(length(fixed_names))

for (i in 1:length(fixed_names))
{
  reduced_formula = as.formula(
    paste("log_area ~",
          paste(fixed_names[-i], collapse = " + "),
          "+ (1 | day) + (1 | month)")
  )
  
  reduced = lmer(reduced_formula, data = df_mixed, REML = FALSE)
  
  LR[i] = -2 * (logLik(reduced)[1] - logLik(full)[1])
  pvalue[i] = pchisq(LR[i], 1, lower.tail = FALSE)
}

fixed_test_table = data.frame(
  Fixed_Effect = fixed_names,
  LR = LR,
  p_value = pvalue
)

fixed_test_table[,2:3] = round(fixed_test_table[,2:3], 4)

print(fixed_test_table)


```
From the likelihood ratio tests, none of the fixed effects are strongly significant at the 0.05 level, since all the p-values are above 0.05. The closest one is DMC, with a p-value around 0.08, which suggests it might have some weak effect, but it’s still not strong enough to be considered statistically significant. ISI and RH also show slightly smaller p-values compared to the others, but they are still not close enough to the cutoff. Overall, this result is pretty consistent with what we’ve been seeing throughout the analysis — it’s really hard to explain variation in fire area using these variables. Most of the predictors don’t contribute much individually, which suggests that the relationship is either very weak or driven by factors not included in the model.

```{r}

VarCorr(full)
sigma = data.frame(summary(full)$varcor)$sdcor
names(sigma) = data.frame(summary(full)$varcor)$grp

ICC = sigma^2 / sum(sigma^2)

print(ICC)
```

```{r}

ICC_table = data.frame(
  Random_Effect = names(ICC),
  ICC = ICC,
  Percent_Total_Variation = 100 * ICC
)

ICC_table[,2:3] = round(ICC_table[,2:3], 4)

print(ICC_table)
```

From the ICC results, we can see that the random effects explain almost none of the total variation in log_area. The month effect only explains about 0.44% of the variation, and day is basically 0%, which means grouping by day of the week doesn’t matter at all here. Almost all the variation, around 99.56%, is coming from the residual, so it’s just unexplained noise. This tells us that fires within the same month are only slightly more similar to each other, and fires on the same day are not similar at all. Another way to interpret the ICC is as the correlation between two observations from the same group, so here it means two fires from the same month have almost no shared similarity, and for the same day, basically none. Overall, the random effects are not contributing much, which again shows that this dataset is really hard to model and most of the variation is coming from factors we didn’t include.

## model selection 

```{r}
set.seed(0)

df_selection = df
df_selection$log_area = log(df_selection$area)

Y = df_selection$log_area

X = df_selection[, c("X", "Y", "FFMC", "DMC", "DC", "ISI", "temp", "RH", "wind")]

names(X)[names(X) == "X"] = "coord_X"
names(X)[names(X) == "Y"] = "coord_Y"

n = dim(X)[1]
P = dim(X)[2]

train = sample(n, 0.6 * n)
valid = sample(setdiff(1:n, train), 0.2 * n)
test = setdiff(1:n, c(train, valid))

MSPE = numeric(P)
```
```{r}
for (p in P:1)
{
  fit = lm(Y ~ ., X[, 1:p, drop = FALSE], train)
  predictions = predict(fit, X[valid, 1:p, drop = FALSE])
  MSPE[p] = mean((Y[valid] - predictions)^2)
  ind = which.max(summary(fit)$coefficients[-1, 4])
  if (ind != p)
  {
    X[, c(ind, p)] = X[, c(p, ind)]
    names(X)[c(ind, p)] = names(X)[c(p, ind)]
  }
}

plot(MSPE,type = "b",pch = 16,lwd = 4,ylab = "Mean Squared Prediction Error")
p = which.min(MSPE)
p

fit = lm(Y ~ ., X[, 1:p, drop = FALSE], test)

summary(fit)
```
Overall, this model selection result is a little disappointing, but it is still informative. The procedure selected ISI because the one-covariate model had the lowest validation MSPE. However, after refitting this selected model on the test set, ISI is clearly not significant. Its p-value is 0.499, and the confidence interval includes zero.

So the main takeaway is that model selection did pick one variable, but valid post-selection inference does not support it as a meaningful predictor. In other words, ISI may look relatively useful during the selection process, but it does not hold up on the test set. The model also has very weak explanatory power, since the adjusted \(R^2\) is negative and the overall F-test is not significant.

This again supports the same conclusion we saw earlier: the burned area variable is difficult to explain with the available covariates. Even after using a model selection method, we still do not get a strong or reliable linear model for \(\log(area)\).

## collinearity  


```{r}

library(glmnet)

X = as.matrix(df_trans[, c("X", "Y", "FFMC", "DMC", "DC", "ISI", "temp", "RH", "wind")])
Y = df_trans$log_area

n = nrow(X)
P = ncol(X)
```

```{r}

## Collinearity check

cor(df_trans[, c("X", "Y", "FFMC", "DMC", "DC", "ISI", "temp", "RH", "wind")])

plot(df_trans[, c("X", "Y", "FFMC", "DMC", "DC", "ISI", "temp", "RH", "wind")],
     pch = 16, cex = 0.25)
```
From the correlation matrix and scatterplot matrix, we can see that some variables show clear relationships with each other, especially among the fire weather variables like DMC, DC, and ISI. The scatterplots show visible linear trends between these variables, which suggests moderate collinearity. This means some predictors are carrying overlapping information, which can make OLS coefficient estimates unstable and sensitive to small changes in the data. So overall, there is some level of multicollinearity present in this dataset.

```{r}

## Train / test split

set.seed(0)

train = sample(1:n, floor(n / 2))
test = (1:n)[-train]
```

```{r}

## Ridge regression

cv.ridge = cv.glmnet(X[train,], Y[train], nfolds = 5, alpha = 0)

ridge = glmnet(X[test,], Y[test],
               alpha = 0,
               lambda = cv.ridge$lambda.min)

print(ridge$beta)
```
From the ridge regression results, we can see that all coefficients are shrunk toward zero compared to OLS, but none of them are exactly zero. This is expected because ridge regression penalizes large coefficients but keeps all variables in the model. The shrinkage helps reduce instability caused by multicollinearity. Compared to OLS, the coefficients are smaller and more balanced, which suggests that ridge is stabilizing the estimates when predictors are correlated.
```{r}

## Bootstrap SE for ridge coefficients

nboot = 1000

beta_ridge = matrix(0, nboot, P)
colnames(beta_ridge) = colnames(X)

for (i in 1:nboot)
{
  ind = sample(test, replace = TRUE)
  beta_ridge[i,] = glmnet(X[ind,], Y[ind],
                          alpha = 0,
                          lambda = cv.ridge$lambda.min)$beta[,]
}

SE_ridge = apply(beta_ridge, 2, sd)

print(SE_ridge)
```
Using bootstrap, we estimated the standard errors for the ridge coefficients. The bootstrap SEs are generally small and more consistent across variables, which indicates that the ridge estimates are relatively stable across different resampled datasets. Since ridge already shrinks coefficients, it reduces variability, so the bootstrap confirms that ridge provides more stable estimates compared to OLS in the presence of collinearity.
```{r}

## Lasso regression

cv.lasso = cv.glmnet(X[train,], Y[train], alpha = 1)

lasso = glmnet(X[test,], Y[test],
               alpha = 1,
               lambda = cv.lasso$lambda.min)

print(lasso$beta)
```
From the lasso regression results, we can see that some coefficients are reduced a lot, and at least one variable (temp) is shrunk all the way to zero. This shows that lasso not only shrinks coefficients but also performs variable selection. Compared to ridge, lasso produces a simpler model by removing less important variables. This makes the model easier to interpret and helps reduce noise from irrelevant predictors.
```{r}
## Bootstrap SE for lasso coefficients

beta_lasso = matrix(0, nboot, P)
colnames(beta_lasso) = colnames(X)

for (i in 1:nboot)
{
  ind = sample(test, replace = TRUE)
  beta_lasso[i,] = glmnet(X[ind,], Y[ind],
                          alpha = 1,
                          lambda = cv.lasso$lambda.min)$beta[,]
}

SE_lasso = apply(beta_lasso, 2, sd)

print(SE_lasso)

```
The bootstrap standard errors for the lasso coefficients show more variation compared to ridge. This is because lasso can set coefficients to zero or near zero, which makes the estimates less stable across resamples. Some variables have larger SEs, indicating that their inclusion in the model is less consistent. Overall, lasso is more aggressive in variable selection, but this can come with higher variability in the estimates.
```{r}

## OLS comparison

ols = lm(log_area ~ X + Y + FFMC + DMC + DC + ISI + temp + RH + wind,
         data = df_trans)

SE_OLS = summary(ols)$coefficients[-1, 2]

comparison_table = data.frame(
  Variable = colnames(X),
  OLS_SE = SE_OLS,
  Ridge_SE = SE_ridge,
  Lasso_SE = SE_lasso
)

comparison_table[,2:4] = round(comparison_table[,2:4], 4)

print(comparison_table)
```

Comparing the standard errors across OLS, ridge, and lasso, we can see that ridge generally produces the smallest standard errors, while OLS tends to have larger ones due to multicollinearity. Lasso standard errors are sometimes larger than ridge because of its variable selection behavior, which introduces more variability. This shows that regularization methods help reduce instability in coefficient estimates, especially when predictors are correlated. However, even with regularization, the differences are not extremely large, which suggests that the overall signal in the data is still weak. So in conclusion, ridge provides the most stable estimates, lasso helps simplify the model, and OLS is more affected by collinearity.

## multiple comparison 

```{r}

## One-way ANOVA (lecture style)
fit_aov = aov(log_area ~ month, data = df_trans)
summary(fit_aov)

## Tukey multiple comparisons (95% global level)
tukey_res = TukeyHSD(fit_aov)
tukey_res
```

We test for differences in the average log_area across all months using a one-way ANOVA and then apply the Tukey multiple comparisons method at the 95% global confidence level to compare all possible month pairs. From the ANOVA result, the p-value is 0.342, which is much larger than 0.05, so we fail to reject the null hypothesis. This means there is no overall evidence that the average fire size differs across months. Looking at the Tukey multiple comparisons results, we see that all adjusted p-values are very large (almost all close to 1), and every confidence interval includes zero. This confirms that none of the month-to-month differences are statistically significant.

Compared to doing pairwise comparisons without any correction, the Tukey method is more reliable because it controls the overall Type I error rate when making many comparisons. Without this adjustment, we could incorrectly find some differences to be significant just due to random chance. In contrast, Tukey’s method shows that once we properly account for multiple testing, there are no meaningful differences between any month pairs. Overall, both the ANOVA and the Tukey results consistently indicate that month does not have a significant effect on log_area, and any observed differences are likely due to random variation rather than real patterns in the data.

## multiple linear model 

```{r}
## Full multiple linear model (all covariates)
fit_full = lm(log_area ~ X + Y + FFMC + DMC + DC + ISI + temp + RH + wind + month + day,
              data = df_trans)

summary(fit_full)

```

```{r}
## ANOVA table (global testing perspective)
anova(fit_full)
```

```{r}
alpha = 0.05

pvalues = summary(fit_full)$coefficients[-1, "Pr(>|t|)"]
m = length(pvalues)

# Lecture 34: Testing Global Null
Bonf = min(pvalues)
print(Bonf < alpha / m)

Fisher = -2 * sum(log(pvalues))
print(Fisher > qchisq(1 - alpha, 2 * m))

Simes = sum(pvalues[order(pvalues)] <= (1:m) * alpha / m)
print(Simes > 0)
```

We fit a full multiple linear regression model using all available quantitative and qualitative covariates to explain log_area. To check whether at least one variable has a significant effect, we use the global F-test from the model summary. The p-value for the F-test is about 0.0805, which is greater than 0.05, so we fail to reject the null hypothesis at the 5% level. This means that overall, the model does not provide strong evidence that at least one predictor is statistically significant, although it is kind of borderline at the 10% level.

Looking at the individual coefficients, we can see that DMC is highly significant and DC is also significant, which suggests that some predictors do have effects on log_area when considered individually. However, the global test is more strict since it tests all variables together, and the overall result is still not strong enough to be considered significant at the 0.05 level.

We also compare this with the ANOVA table, which provides another way to assess significance. The ANOVA results show that the variable month is statistically significant (p = 0.0277), while most other variables are not. So unlike the global F-test, the ANOVA suggests that at least one variable (month) does have a significant effect.

Overall, the two methods give slightly different conclusions. The global F-test suggests no strong overall significance, while the ANOVA highlights that certain variables like month may still be important. This difference happens because the global test evaluates all variables together, while ANOVA looks at each variable’s contribution separately.

**add 3 more tests, this write up need fix!** The Bonferroni global test rejects the global null hypothesis because the smallest p-value is smaller than alpha divided by the number of tests. The Simes test also rejects the global null hypothesis, since at least one ordered p-value is smaller than its corresponding Simes threshold. However, Fisher's combination test does not reject the global null hypothesis.

These results do not fully agree because the three methods aggregate evidence differently. Bonferroni and Simes can reject the global null when there is at least one sufficiently small p-value. Fisher's method combines evidence across all p-values, so it is more sensitive when several effects show moderate evidence at the same time. In our case, the results suggest that the evidence against the global null is mainly driven by one or a small number of strong individual effects, rather than by broad evidence across many effects.


## Multiple testing methods 
```{r}
## Part 14: Multiple testing corrections

alpha = 0.05

fit_full = lm(log_area ~ X + Y + FFMC + DMC + DC + ISI + temp + RH + wind + month + day,
              data = df_trans)

pvalues = summary(fit_full)$coefficients[-1, 4]
m = length(pvalues)

## No correction
No_Correction = pvalues

## Holm-Bonferroni (FWER control)
HB = p.adjust(pvalues, "holm")

## Benjamini-Hochberg (FDR control)
BH = p.adjust(pvalues, "BH")

## Benjamini-Yekutieli procedure
BY = p.adjust(pvalues, "BY")
```


```{r}
testing_table = data.frame(
  Effect = names(pvalues),
  No_Correction = No_Correction,
  Holm_Bonferroni = HB,
  Benjamini_Hochberg = BH,
  Benjamini_Yekutieli = BY
)

testing_table[,2:5] = round(testing_table[,2:5], 4)

print(testing_table)
```
From the table, we can see how the significance of variables changes once we apply multiple testing corrections. Without any correction, both DMC (p = 0.0009) and DC (p = 0.0137) look statistically significant at the 5% level. However, after applying the Holm-Bonferroni method, only DMC remains significant (p = 0.0224), while DC is no longer significant. The Benjamini-Hochberg (BH) method gives a similar result, where DMC is still significant, but DC again becomes non-significant. All other variables have large p-values across all methods, so they are clearly not significant. Overall, Holm is more strict since it controls the chance of any false positives, while BH is a bit more flexible, but in this case both methods agree that only DMC has a reliable effect on the response. This suggests that most predictors do not have strong evidence of association with log_area, and the signal in this dataset is pretty weak overall.
```{r}
## Clean table

sum(No_Correction < alpha)
sum(HB < alpha)
sum(BH < alpha)
sum(BY < alpha)
```

From the results, we can see that without any correction there are 2 variables that appear statistically significant at the 5% level. However, after applying the Holm-Bonferroni method, only 1 variable remains significant, and the same happens for the Benjamini-Hochberg method. This shows that once we control for multiple testing, some of the originally “significant” results were likely false positives. The Holm method is more conservative since it controls the probability of making any Type I error, while BH controls the proportion of false discoveries, but in this case both methods agree and reduce the number of significant variables to just one. Overall, this suggests that the evidence for most predictors is weak, and only one variable has a reliable effect on the response after accounting for multiple comparisons.

## Final Model 

Taking all the previous analyses together, the final model should be based on the log-transformed response (log_area) since earlier diagnostic plots showed strong right skewness and the transformation improved residual behavior. From the mixed-effects model, the ICC results showed that month explains only about 0.44% of the variation and day essentially 0%, so random effects do not contribute meaningfully and should be excluded from the final model.

For model selection, the AIC procedure led to a reduced model, and the final fitted model included DMC, ISI, RH, and wind. However, from the regression output, only ISI showed marginal significance (p ≈ 0.014), while others were not significant. In addition, regularization methods (ridge and lasso) showed that most coefficients shrink toward zero, and after multiple testing corrections (Holm and BH), only DMC remained significant, while all other variables became non-significant. This suggests that DMC is the most stable and reliable predictor, while the rest have weak or unstable effects.

Collinearity was observed in the scatterplot matrix (especially among variables like DMC, DC, and ISI), which justifies the use of ridge and lasso regression to stabilize estimates. Since heteroscedasticity was not a major issue based on earlier tests, we do not need WLS, but to ensure valid inference, we should still use robust (HC) standard errors. Because the data is highly skewed and the model assumptions are weak, inference should also be supported using bootstrap standard errors and confidence intervals, instead of relying only on asymptotic results.

For hypothesis testing, we used global F-tests, likelihood-based methods, and permutation-style reasoning earlier, and results consistently showed weak overall significance (e.g., global p-value ≈ 0.08). Multiple testing corrections further confirmed that most variables are not significant once we control for false positives.

Overall, the final model should be a log-linear model with a small set of predictors (primarily DMC, with cautious inclusion of ISI), no random effects, and inference based on robust SEs, bootstrap methods, and multiple testing corrections. Even after all adjustments, the model has very low explanatory power (adjusted R² ≈ 0.016), meaning that most of the variation in fire size is still unexplained (only explain 1.6% of variation in log_area), likely due to missing important variables or the inherently unpredictable nature of wildfire behavior.

